---
title: "ML cancer"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
knitr::opts_chunk$set(fig.width=12, fig.height=8) 
library(ggplot2)
library(reshape2)
library(caret)
library(corrplot)
library(RColorBrewer)
library(data.table)
library(gridExtra)
library(randomForest)
library(png)
load("cancer.RData")
```

## 1

### a) Var types

```{r}
var_types <- sapply(data.train, class)
print(var_types[!duplicated(var_types)])
```

```
ENSG00000000003 
      "numeric"
```

### b) Correlation 500 most variability genes

```{r}
vars <- sapply(within(data.train, rm(Y)), var)
vars_df <- data.frame(vars)
vars_test <- sapply(data.test, var)
vars_test_df <- data.frame(vars_test)
vars500 <- sort(vars, decreasing = TRUE)[1:500]
vars500_data <- data.train[,names(vars500)]
vars500_data_all <- vars500_data
vars500_data_all$Y <- data.train$Y
cors500 <- cor(vars500_data)
cors500[lower.tri(cors500, diag = TRUE)] <- NA
melted_cors500 <- melt(cors500)
cors500_final <- na.omit(melted_cors500)
```

```{r, echo=FALSE}
ggplot(cors500_final, aes(x=value, y="")) + geom_violin() + ylab("") + xlab("Correlation") + ggtitle("Correlations distribution for 500 most varianced genes")
```

```{r, echo=FALSE, eval=TRUE, fig.height=8, fig.width=12, fig.align="center"}
pp1 <- readPNG("zad1.png")
plot(NA, xlim = c(0, 1), ylim = c(0, 1), type = "n", xaxt = "n", yaxt = "n", xlab = "", ylab = "")
rasterImage(pp1,0,0,1,1)
```

## 3

Histograms:

```{r, echo=FALSE}
ggplot(vars_df, aes(vars)) + geom_histogram(binwidth = 0.1, colour = "black")
ggplot(vars_test_df, aes(vars_test)) + geom_histogram(binwidth = 0.1, colour = "black")
```


```{r, echo=FALSE, eval=TRUE, fig.height=8, fig.width=12, fig.align="center"}
pp1 <- readPNG("zad2_1.png")
plot(NA, xlim = c(0, 1), ylim = c(0, 1), type = "n", xaxt = "n", yaxt = "n", xlab = "", ylab = "")
rasterImage(pp1,0,0,1,1)
```

```{r, echo=FALSE, eval=TRUE, fig.height=8, fig.width=12, fig.align="center"}
pp1 <- readPNG("zad2_2.png")
plot(NA, xlim = c(0, 1), ylim = c(0, 1), type = "n", xaxt = "n", yaxt = "n", xlab = "", ylab = "")
rasterImage(pp1,0,0,1,1)
```

ANOVA selection:

```{r}
anova_threshold <- 0.001
data_to_train <- data.train
l <- vector("list", ncol(data_to_train) - 1)
j <- 1
for (i in 1:(ncol(data_to_train) - 1)) {
  gene <- names(data_to_train[i])
  anovaresult <- anova(aov(data_to_train[,i] ~ Y,data=data_to_train))
  if(anovaresult$Pr[1] < anova_threshold) {
    p <- anovaresult$Pr[1]
    l[[j]] <- gene
    j <- j + 1
  }
}

an_name <- paste("anova_threshold_", anova_threshold, sep = "")
an_name <- sub("-", "", an_name)
assign(an_name, l[lengths(l) != 0])

nl <- eval(parse(text = an_name))
print(paste("ANOVA genes count: ", length(nl), sep=""))

genes_after_anova <- unlist(nl, use.names = FALSE)
genes_after_anova_with_Y <- append(genes_after_anova, 'Y')
data_after_anova <- data.train[,genes_after_anova_with_Y]
data_after_anova_raw <- data.train[,genes_after_anova]

cdata <- data_after_anova #  data_after_corr
cory <- abs(cor(cdata))

cory_y <- data.frame(cory[,"Y"])
names(cory_y)[1] <- "y_cor"
melt(setDT(cory_y, keep.rownames = TRUE))
```
```
[1] "ANOVA genes count: 7253"
```

Correlation selection:

```{r}
find_best <- function(corr_threshold, btw_corr_threshold, mode, data_train, data_test) {
  res <- list()
  
  res$corr_threshold = corr_threshold
  res$btw_corr_threshold = btw_corr_threshold
  
  if (corr_threshold >= 0) {
    # Y correlation
    cory_y_sel <- cory_y[cory_y$y_cor > corr_threshold & cory_y$y_cor != 1.0,]
    cor_y_names <- as.vector(cory_y_sel[,1]$rn)
    data_after_corr_y <- cdata[,cor_y_names]
  } else { # skip if -1
    data_after_corr_y <- cdata
  }
  
  if (btw_corr_threshold >= 0) {
    # between correlation
    cdata <- data_after_corr_y# data_after_anova_raw
    hc = findCorrelation(cor(cdata), cutoff = btw_corr_threshold)
    hc = sort(hc)
    data_after_corr <- cdata[,-c(hc)]
    data_after_corr$Y <- data.train$Y
  } else { # skip if -1
    data_after_corr <- cdata
    data_after_corr$Y <- data.train$Y
  }
  
  cdata <- data_after_corr  # data_after_corr_y
  cgenes <- paste(unlist(colnames(cdata)[1:ncol(cdata) - 1]), collapse=' + ')
  
  trnCtrl <- trainControl(method = "repeatedcv", number = 10, repeats = 3)
  #trnCtrl <- trainControl(method = "cv", number = 10)
  
  if (mode == "glmnet") {
    srchGrd <- expand.grid(alpha = seq(0, 1, length = 10), lambda = seq(0.0001, 1, length = 100))
    mod_train <- paste("my.train5 <- train(Y ~", cgenes, ", data = data_train, method = \"glmnet\", tuneGrid = srchGrd, trControl = trnCtrl, standardize = TRUE, maxit = 10000)")
  }
  else if (mode == "rf") {
    rfGrd <- expand.grid(.mtry=c(sqrt(ncol(cdata))))
    mod_train <- paste("my.train5 <- train(Y ~", cgenes, ", data = data_train, method = \"rf\", tuneGrid = rfGrd, trControl = trnCtrl, standardize = TRUE, maxit = 10000)")
  }
  else {
    return(res)
  }
  
  eval(parse(text = mod_train))
  
  res$model = my.train5
  res$mod_train = mod_train
  
  # cross validated train RMSE
  if (mode == "glmnet") {
    res$train_err = my.train5$model$results$RMSE[which.min(my.train5$model$results$RMSE)]
  }
  else if (mode == "rf") {
    res$train_err = my.train5$model$results$RMSE
  }
  
  # test RMSE
  preds_en5_test <- predict(my.train5, newdata = data_test)
  res$test_err = postResample(preds_en5_test, data_test$Y)
  
  res
}
```

```{r}
find_best_btw_corr <- function(min_y_corr, btw_corr_threshold, mode, final_data) {
  if (final_data == TRUE) { # final data to train on
    a_train = data.train
    a_test = data.train
  }
  else { # splitted training data to find out best model
    a_train <- data.train[100:644,]
    a_test <- data.train[1:100,]
  }
  set.seed(2137)
  
  results = list()
  j = 1
  if (min_y_corr >= 0) {
    for (i in seq(0.5, min_y_corr, length.out = 11)) {
      res <-find_best(i, btw_corr_threshold, mode, a_train, a_test)
      results[[j]] = res
      j = j + 1
      print(res$corr_threshold)
      print(length(res$model$coefnames))
      print(res$train_err)
      print(res$test_err)
    }
  }
  else { # when min y corr is <0 than skip selecting based on it
    res <-find_best(-1, btw_corr_threshold, mode, a_train, a_test)
    results[[j]] = res
    print(res$corr_threshold)
    print(length(res$model$coefnames))
    print(res$train_err)
    print(res$test_err)
  }
  results
}
```

Results:

```{r}
show_model_results_plot <- function(results, title, filename) {
  datalist = list()
  j = 1
  for (res in results) {
    datalist[[j]] = data.frame(corr_y=res$corr_threshold, test_rmse=res$test_err[1], train_rmse=res$train_err, num=length(res$model$coefnames))
    j = j + 1
  }
  df = do.call(rbind, datalist)
  rownames(df) <- NULL
  
  p1 = ggplot(df, aes(x=corr_y)) + geom_line(aes(y=test_rmse), color='red') + geom_line(aes(y=train_rmse), color='blue') + labs(y = 'RMSE', x = 'Y correlation') + ylim(0.1, 0.14) + ggtitle(title)
  p2 = ggplot(df, aes(x=num)) + geom_line(aes(y=test_rmse), color='red') + geom_line(aes(y=train_rmse), color='blue') + labs(y = 'RMSE', x = 'Number of genes') + ylim(0.1, 0.14) + xlim(0, 7253) + ggtitle(title)
  plts = grid.arrange(p1, p2, nrow = 2)
  ggsave(filename, plot=plts)
}
```

1. ANOVA selection (based on p-value threshold)
2. Select variables most correlated with 'Y' (based on Y-corr threshold)
3. Removing most cross-correlated variables (based on cross-corr threshold)

## 4

Building models:

```{r}
res_glmnet_065 = find_best_btw_corr(0.1, 0.65, "glmnet", FALSE)
results_glmnet_090 = find_best_btw_corr(0.1, 0.9, "glmnet", FALSE)
res_rf_065 = find_best_btw_corr(0.1, 0.65, "rf", FALSE)
results_rf_090 = find_best_btw_corr(0.1, 0.9, "rf", FALSE)
```

Models need to be saved and read after built, due to memory limits:

```{r, echo=FALSE}
res_glmnet_065 = readRDS("res_glmnet_065.rds")
res_glmnet_090 = readRDS("res_glmnet_090.rds")
res_rf_065 = readRDS("res_rf_065.rds")
res_rf_090 = readRDS("res_rf_090.rds")
```

Results:

```{r, echo=FALSE}
show_model_results_plot(res_glmnet_065, "Elastic net, cutoff correlation between = 0.65", "glmnet_065.png")
show_model_results_plot(res_glmnet_090, "Elastic net, cutoff correlation between = 0.90", "glmnet_090.png")
show_model_results_plot(res_rf_065, "Random forest, cutoff correlation between = 0.65", "rf_065.png")
show_model_results_plot(res_rf_090, "Random forest, cutoff correlation between = 0.90", "rf_090.png")
```

```{r, echo=FALSE, eval=TRUE, fig.height=8, fig.width=12, fig.align="center"}
pp1 <- readPNG("glmnet_065.png")
pp2 <- readPNG("glmnet_090.png")
pp3 <- readPNG("rf_065.png")
pp4 <- readPNG("rf_090.png")
plot(NA, xlim = c(0, 2), ylim = c(0, 2), type = "n", xaxt = "n", yaxt = "n", xlab = "", ylab = "")
rasterImage(pp1,0,0,1,1)
rasterImage(pp2,0,1,1,2)
rasterImage(pp3,1,0,2,1)
rasterImage(pp4,1,1,2,2)
```

Best model is Random Forest with Y-correlation threshold = 0.46 and cross-correlation threshold = 0.9 on data with ANOVA p-value threshold = 0.001

-------

Models without any selection:

```{r}
res_glmnet_000 = find_best_btw_corr(-1, -1, "glmnet", FALSE)
res_rf_000 = find_best_btw_corr(-1, -1, "rf", FALSE)
```


```{r, echo=FALSE}
print(paste("elastic net (pure) RMSE: ", res_glmnet_000[[1]]$train_err[[1]], "; random forest (corr_y = 0.46, corr_between = 0.9) RMSE: ", res_rf_090[[2]]$train_err[[1]]))
```

```
elastic net   (pure)                              RMSE:  0.102322042198110
random forest (corr_y = 0.46, corr_between = 0.9) RMSE:  0.106720089796651
```

```{r, echo=FALSE}
print(paste("elastic net (pure) RMSE: ", res_glmnet_000[[1]]$test_err[[1]], "; random forest (corr_y = 0.46, corr_between = 0.9) RMSE: ", res_rf_090[[2]]$test_err[[1]]))
```

```
elastic net   (pure)                              RMSE:  0.132425543474816
random forest (corr_y = 0.46, corr_between = 0.9) RMSE:  0.121851519350148
```

Best Elastic Net models has also Y-correlation threshold = 0.46 and cross-correlation threshold = 0.9 on data with ANOVA p-value threshold = 0.001

```{r, echo=FALSE}
print(paste("elastic net (corr_y = 0.46, corr_between = 0.9) RMSE: ", res_glmnet_090[[2]]$train_err[[1]], "; random forest (corr_y = 0.46, corr_between = 0.9) RMSE: ", res_rf_090[[2]]$train_err[[1]]))
```

```
elastic net   (corr_y = 0.46, corr_between = 0.9) RMSE:  0.105508153223289
random forest (corr_y = 0.46, corr_between = 0.9) RMSE:  0.106720089796651
```

```{r, echo=FALSE}
print(paste("elastic net (corr_y = 0.46, corr_between = 0.9) RMSE: ", res_glmnet_090[[2]]$test_err[[1]], "; random forest (corr_y = 0.46, corr_between = 0.9) RMSE: ", res_rf_090[[2]]$test_err[[1]]))
```

```
elastic net   (corr_y = 0.46, corr_between = 0.9) RMSE:  0.123187638354296
random forest (corr_y = 0.46, corr_between = 0.9) RMSE:  0.121851519350148
```

Random Forest without any selections:

```{r, echo=FALSE}
print(paste("random forest (pure) RMSE: ", res_rf_000[[1]]$train_err[[1]], "; random forest (corr_y = 0.46, corr_between = 0.9) RMSE: ", res_rf_090[[2]]$train_err[[1]]))
```

```
random forest (pure)                              RMSE:  0.104863668843563
random forest (corr_y = 0.46, corr_between = 0.9) RMSE:  0.106720089796651
```

```{r, echo=FALSE}
print(paste("random forest (pure) RMSE: ", res_rf_000[[1]]$test_err[[1]], "; random forest (corr_y = 0.46, corr_between = 0.9) RMSE: ", res_rf_090[[2]]$test_err[[1]]))
```

```
random forest (pure)                              RMSE:  0.124790824997018
random forest (corr_y = 0.46, corr_between = 0.9) RMSE:  0.121851519350148
```

Final results:

```
Method                                                  RMSE train          RMSE test
---------------------------------------------------------------------------------------------
elastic net (pure):                                     0.102322042198110   0.132425543474816
elastic net (corr_y = 0.46, corr_between = 0.9):        0.105508153223289   0.123187638354296
random forest (pure):                                   0.104863668843563   0.124790824997018
random forest (corr_y = 0.46, corr_between = 0.9):      0.106720089796651   0.121851519350148
```

## 5

Selected genes for best model (Random Forest Y-corr = 0.46; cross-corr = 0.9)

```
 [1] "ENSG00000006327" "ENSG00000007968" "ENSG00000061676" "ENSG00000068831" "ENSG00000072682" "ENSG00000075884" "ENSG00000082781"
 [8] "ENSG00000085733" "ENSG00000089094" "ENSG00000089876" "ENSG00000100196" "ENSG00000102879" "ENSG00000104067" "ENSG00000105122"
[15] "ENSG00000107099" "ENSG00000118308" "ENSG00000118513" "ENSG00000122122" "ENSG00000123338" "ENSG00000124145" "ENSG00000125354"
[22] "ENSG00000130755" "ENSG00000130958" "ENSG00000134516" "ENSG00000136830" "ENSG00000137693" "ENSG00000138119" "ENSG00000143851"
[29] "ENSG00000160633" "ENSG00000160789" "ENSG00000162909" "ENSG00000167984" "ENSG00000168421" "ENSG00000168938" "ENSG00000172380"
[36] "ENSG00000180448" "ENSG00000181019" "ENSG00000182162" "ENSG00000186517" "ENSG00000189171" "ENSG00000196247" "ENSG00000198771"
[43] "ENSG00000205730"
```

Building final model with cross-validation, tuning ntree:

```{r}
res_rf_090 = readRDS("res_rf_090.rds")
genes = res_rf_090[[2]][["model"]][["coefnames"]]
whole_data_train = data.train[,genes]
whole_data_train$Y = data.train$Y

trnCtrl <- trainControl(method = "cv", number = 10)
rfGrd <- expand.grid(.mtry=c(sqrt(ncol(cdata))))

modellist <- list()
for (ntree in c(1000,1500,2000,2500)){
  set.seed(2137)
  final_model = train(Y ~ ., data = whole_data_train, method = "rf", tuneGrid = rfGrd, trControl = trnCtrl, standardize = TRUE, maxit = 10000)
  modellist[[toString(ntree)]] = final_model
  print(final_model)
}


print(final_model)
```
```
Random Forest 

644 samples
 43 predictor

No pre-processing
Resampling: Cross-Validated (10 fold, repeated 5 times) 
Summary of sample sizes: 580, 580, 580, 580, 580, 580, ... 
Resampling results:

  RMSE       Rsquared  MAE       
  0.1088438  0.383584  0.08162398

Tuning parameter 'mtry' was held constant at a value of 6.63325
```

Prediction:

```{r}
pred <- predict(final_model, newdata = data.test)
save(pred, file="pred_results.RData")
```